Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-12-20 article.tex
Explain B
More code and some sven notes

@@

Another approach, which is a much more precise approximation, is a specific variant of the dataflow analysis as described in \textit{Modern Compiler Implementation in ML} (\cite{tiger}) and \textit{Compilers: Principles, Techniques, and Tools} (\cite{dragon}). In general, dataflow analysis is the process of finding the possible paths in which data may propagate through different branches of execution.  While several applications of this exist (like constant propagation, reaching defnitions, available expressions etc.), one that is immediately beneficial in the case of liveness analysis is one that traverses a CFG in the reverse order of execution (i.e.  \textit{backwards} flow), and extracts any variable that \textit{may} be used in execution (also referred to as \textit{backwards may} analysis).

@@

\subsubsection{Implementation}

The \texttt{dataflow} function takes a list of instructions \texttt{insns}, an array of CFG vertices \texttt{ids} as well as the CFG \texttt{g}. The \(i\)th element of \texttt{ids} is the CFG vertex corresponding to the \(i\)th entry of \texttt{insns}. The decision to use arrays instead of pairing up each instruction with both \textit{live-in} and \textit{live-out} sets was made because of the \textit{live-out} equation \eqref{flowout} that needs to find all \textit{live-out} sets of the successor blocks. Hence random access arrays were chosen instead of linear search.
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let dataflow (insns : Cfg.insn list) (ids : Cfg.G.V.t array) (g : Cfg.G.t) =
  let insns = List.mapi (fun i v -> (i, v)) insns |> List.rev in
  let in_ = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let out = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let rec dataflow () =
    let flowout = (* ... *)
    let flowin = (* ... *)
    let flow changed insn = changed || flowout insn || flowin insn in
    if List.fold_left flow false insns then dataflow () else (in_, out)
  in
  dataflow ()
\end{minted}
The \texttt{flowin} function correponds to the \textit{live-in} equation \eqref{flowin} and is implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowin (i, insn) =
  let newin = S.SS.union (use insn) (S.SS.diff out.(i) (def insn)) in
  let changed = not (S.SS.equal newin in_.(i)) in
  if changed then in_.(i) <- newin;
  changed
\end{minted}
Where \lstinline!S.SS! is a \lstinline!Set.S! module built over the \lstinline!symbol! type found in \lstinline!lib/symbol.ml!:

\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
type symbol = string * int
(* ... *)
module SS = Set.Make (struct
  type t = symbol
  let compare (_, n1) (_, n2) = compare n1 n2
end)
type set = SS.t
\end{minted}

\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowout (i, _) =
  let newout =
    let succ = Cfg.G.succ g ids.(i) in
    List.fold_left
      (fun s v -> S.SS.union s in_.(Cfg.G.V.label v))
      S.SS.empty succ
  in
  let changed = not (S.SS.equal newout out.(i)) in
  if changed then out.(i) <- newout;
  changed
\end{minted}

@@

Because of this, for any phi node of the form 
\[
  a = \Phi (b_1, ..., b_n)
\]
add edges to all \textit{live-out} variables not in \(B\)
\[
  \forall b_j \in (\mathit{out}[i] \setminus B), \mathit{add\_edge}(a, b_j)
\]
For any other instruction the defines a variable \(a\)
\[
  \forall b_j \in B, \mathit{add\_edge}(a, b_j)
\]
Although the Appel text notes interference with concrete registers as well as overlapping variables, this isn't considered in this implementation.

>>>>

% FIXME: introducer phi nodes og nævn dominators i CFA afsnittet

@@

% FIXME: beskriv use, def osv. sets tidligere!

Another approach, which is a much more precise approximation, is a specific variant of the dataflow analysis as described in \textit{Modern Compiler Implementation in ML} (\cite{tiger}) and \textit{Compilers: Principles, Techniques, and Tools} (\cite{dragon}). In general, dataflow analysis is the process of finding the possible paths in which data may propagate through different branches of execution.  While several applications of this exist (like constant propagation, reaching defnitions, available expressions etc.), one that is immediately beneficial in the case of liveness analysis is one that traverses a CFG in the reverse order of execution (i.e.  \textit{backwards} flow), and extracts any variable that \textit{may} be used in execution (also referred to as \textit{backwards may} analysis). % FIXME: forklar hvorfor List.rev ikke udgør backwards flow

@@

%\subsubsection{Implementation}

The simplest ones to implement are the \lstinline!def! and \lstinline!use! functions, as all of the values of interest are located immediately within the instruction itself and not hidden behind some layer of indirection:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let def (s : S.SS.t) (insn : Cfg.insn) =
  match insn with Insn (Some dop, _) -> S.SS.add dop s | _ -> s

let use (s : S.SS.t) (insn : Cfg.insn) =
  let op o s = match o with Ll.Id i -> S.SS.add i s | _ -> s in
  let po s o = op o s in
  match insn with
  | Insn (_, AllocaN (_, (_, o)))
  | Insn (_, Bitcast (_, o, _))
  | Insn (_, Load (_, o))
  | Insn (_, Ptrtoint (_, o, _))
  | Insn (_, Sext (_, o, _))
  | Insn (_, Trunc (_, o, _))
  | Insn (_, Zext (_, o, _)) ->
      op o s
  | Insn (_, Binop (_, _, l, r))
  | Insn (_, Icmp (_, _, l, r))
  | Insn (_, Store (_, l, r)) ->
      op l s |> op r
  | Insn (_, Call (_, _, args)) -> List.map snd args |> List.fold_left po s
  | Insn (_, Gep (_, bop, ops)) -> List.fold_left po (op bop s) ops
  | Insn (_, Select (c, (_, l), (_, r))) -> op c s |> op l |> op r
  | Insn (_, PhiNode (_, ops)) -> List.map fst ops |> List.fold_left po s
  | Term (Ret (_, Some o) | Cbr (o, _, _)) -> op o s
  | _ -> s
\end{minted}
Where \lstinline!S.SS! is a \lstinline!Set.S! module built over the \lstinline!symbol! type found in \lstinline!lib/symbol.ml!:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
type symbol = string * int
(* ... *)
module SS = Set.Make (struct
  type t = symbol
  let compare (_, n1) (_, n2) = compare n1 n2
end)
type set = SS.t
\end{minted}
With the actual fixed-point iteration performed as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let dataflow (insns : Cfg.insn list) (ids : Cfg.G.V.t array) (g : Cfg.G.t) =
  let insns = List.mapi (fun i v -> (i, v)) insns |> List.rev in
  let in_ = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let out = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let rec dataflow () =
    let flowout = (* ... *)
    let flowin = (* ... *)
    let flow changed insn = changed || flowout insn || flowin insn in
    if List.fold_left flow false insns then dataflow () else (in_, out)
  in
  dataflow ()
\end{minted}
The \texttt{flowin} function correponds to the \textit{live-in} equation \eqref{flowin} and is implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowin (i, insn) =
  let newin = S.SS.union (use insn) (S.SS.diff out.(i) (def insn)) in
  let changed = not (S.SS.equal newin in_.(i)) in
  if changed then in_.(i) <- newin;
  changed
\end{minted}
And the \texttt{flowout} function which correponds to the \textit{live-out} equation \eqref{flowout} implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowout (i, _) =
  let newout =
    let succ = Cfg.G.succ g ids.(i) in
    List.fold_left
      (fun s v -> S.SS.union s in_.(Cfg.G.V.label v))
      S.SS.empty succ
  in
  let changed = not (S.SS.equal newout out.(i)) in
  if changed then out.(i) <- newout;
  changed
\end{minted}

@@

Because of this, for any phi node of the form 
\[
  a = \Phi (b_1, ..., b_n)
\]
add edges to all \textit{live-out} variables not in \(B\)
\[
  \forall b_j \in (\mathit{out}[i] \setminus B), \mathit{add\_edge}(a, b_j) \text{ where } B = \{b_1, ..., b_n\}
\]
For any other instruction the defines a variable \(a\)
\[
  \forall b_j \in B, \mathit{add\_edge}(a, b_j)
\]
Although the Appel text notes interference with concrete registers as well as overlapping variables, this isn't considered in this implementation.
